---
title: "Advent of code 2022: day 1"
author: "Quentin Leclerc"
date: "01/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2022/day/1

In this first challenge, the aim is essentially to break up a vector into smaller vectors (using empty spaces
as the separators), sum up each small vector, and return the largest sum.

... However, I completely failed to identify how to break up the vector (maybe using 'split()' or 'cut()'??)
so instead I used a very ugly but straightforward loop :(

Oh well, if it works, it's still a valid solution I guess!

## Solution


```{r test_solution1}

input = as.numeric(readLines(here::here("2022", "inputs", "input1.txt")))

s = 0 #for current chunk sum
tot = 0 #to store all chunk sums

#keep adding to chunk, except if the value is NA, in which case move to the next chunk
for(i in 1:length(input)){
  if(!is.na(input[i])) s = s+input[i]
  else{
    tot = c(tot, s)
    s = 0
  }
}

#return the max sum of chunks
max(tot)

```



# Part 2

## Challenge

The second part can be easily obtained from the first, since I worked out all the sums.
Here, I just sort all the sums, take the 3 biggest, and add them up.

## Solution

```{r result2}

sum(tail(sort(tot),3))

```

