---
title: "Advent of code 2022: day 2"
author: "Quentin Leclerc"
date: "02/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

Today, we're playing rock-paper-scissors!
So, let's figure out how to use the information provided to *definitely not cheat*...
Long stoy short, my solution here is to convert letters to numbers, then work out the difference between my
opponent's and my numbers, and use that to figure out who wins.

Here I tried to avoid using a loop, so instead I applied a few conditional replacements to get my solution.
I'm also trying to stick to base R, so ended up using the base pipe `|>` and nothing from `dplyr` like `mutate()`,
which would have made my life way easier.

## Solution


```{r result1}

input = readLines(here::here("2022", "inputs", "input2.txt")) |>
  strsplit(" ") |>
  unlist() |>
  matrix(ncol = 2, byrow = T)

input = gsub("X", "A", input)
input = gsub("Y", "B", input)
input = gsub("Z", "C", input)

input = gsub("A", "1", input)
input = gsub("B", "2", input)
input = gsub("C", "3", input)

input = apply(input, c(1,2), as.numeric)

input = cbind(input, input[,1]-input[,2])

input[input[,3]==0, 3] = 3
input[input[,3]==2, 3] = 6
input[input[,3]==1, 3] = 0
input[input[,3]==-1, 3] = 6
input[input[,3]==-2, 3] = 0

sum(input[,c(2,3)])

```




# Part 2

## Challenge

Part 2 is the same logic, but flipped problem: we now have the outcome (win, lose, draw),
and have to figure out the input (rock, paper, scissors).
So I'm again using a similar logic with some conditional reformatting, although I feel there must be something 
cleaner to do with `%%`, hmm...

## Solution


```{r result2}

input = readLines(here::here("2022", "inputs", "input2.txt")) |>
  strsplit(" ") |>
  unlist() |>
  matrix(ncol = 2, byrow = T)

input = gsub("X", "0", input)
input = gsub("Y", "3", input)
input = gsub("Z", "6", input)

input = gsub("A", "1", input)
input = gsub("B", "2", input)
input = gsub("C", "3", input)

input = apply(input, c(1,2), as.numeric)

input = cbind(input, 0)

input[input[,2]==3, 3] = input[input[,2]==3, 1]

input[input[,2]==6, 3] = input[input[,2]==6, 1]+1
input[input[,3]==4, 3] = 1

input[input[,2]==0, 3] = input[input[,2]==0, 1]-1
input[input[,3]==0, 3] = 3

sum(input[,c(2,3)])


```

