---
title: "Advent of code 2022: day 3"
author: "Quentin Leclerc"
date: "03/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

Today's challenge is about finding matching characters, so I suppose the solution
has something to do with `match()`.

Unfortunately, I couldn't figure out how to do this without a loop... Maybe some sort
of `apply()` approach would have worked, but here speed was not a problem so I just
went with the loop.

The idea is straightforward: for each line (ie each bag), split it into two compartments,
then search for the matching character between the compartments (using `pmatch()` to only
compare once and go faster). Last step is to convert the matching letter to a value,
for which I can conveniently use the `letters` and `LETTERS` objects included in base `R`.

## Solution


```{r result1}

input = readLines(here::here("2022", "inputs", "input3.txt"))

tot = 0

for(i in 1:length(input)){
  input_i = unlist(strsplit(input[i], ""))
  i1 = input_i[1:(length(input_i)/2)]
  i2 = input_i[(length(input_i)/2+1):length(input_i)]
  
  input_i = i2[pmatch(i1,i2)]
  input_i = input_i[!is.na(input_i)][1]
  
  input_i = match(input_i, c(letters,LETTERS))
  tot = tot+input_i
}

tot

```




# Part 2

## Challenge

Exactly the same problem as Part 1 (finding a unique, matching character), but now
we're comparing 3 bags instead of 2 compartments. So, just need to add one extra step
to the loop, first comparing bags 1 and 2, then comparing this comparison with bag 3.


## Solution


```{r result2}

input = readLines(here::here("2022", "inputs", "input3.txt"))

tot = 0

for(i in seq(1, length(input), 3)){
  input_1 = unlist(strsplit(input[i], ""))
  input_2 = unlist(strsplit(input[i+1], ""))
  input_3 = unlist(strsplit(input[i+2], ""))
  
  input_i = input_2[pmatch(input_1,input_2)]
  input_i = input_i[!is.na(input_i)]
  
  input_i = input_3[pmatch(input_i,input_3)]
  input_i = input_i[!is.na(input_i)][1]
  
  input_i = match(input_i, c(letters,LETTERS))
  tot = tot+input_i
}

tot

```

