---
title: "Advent of code 2022: day 4"
author: "Quentin Leclerc"
date: "04/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

Straightforward challenge today (feels like the difficulty is a bit of a rollercoaster this year??),
we have to figure out if, given two sequences, either is contained fully into the other.
The input is conveniently formatted, so a bit of `gsub()` to remove all non-numerics does the trick
to then extract the numbers defining the sequences. Once that's done, just need to check if either is
contained in the other using `is.element()` and `all()`.

## Solution


```{r result1}

input = readLines(here::here("2022","inputs","input4.txt"))

n_pairs = 0

for(i in input){
  
  i = gsub("\\D", " ", i) |>
    strsplit(" ") |>
    unlist() |>
    as.numeric()
  
  if(all(is.element(c(i[1]:i[2]),c(i[3]:i[4]))) || all(is.element(c(i[3]:i[4]),c(i[1]:i[2])))){
    n_pairs = n_pairs+1
  }

}

n_pairs

```







# Part 2

## Challenge

Now we want to know if there's any overlap at all between sequences, which is arguably
easier than Part 1. Same logic as above for the formatting, but now we can just
use `intersect()` to see if the sequences have anything in common!

## Solution


```{r result2}

input = readLines(here::here("2022","inputs","input4.txt"))

n_pairs = 0

for(i in input){
  
  i = gsub("\\D", " ", i) |>
    strsplit(" ") |>
    unlist() |>
    as.numeric()
  
  if(length(intersect(c(i[1]:i[2]), c(i[3]:i[4]))) > 0){
    n_pairs = n_pairs+1
  }

}

n_pairs

```

