---
title: "Advent of code 2022: day 5"
author: "Quentin Leclerc"
date: "05/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

A frustrating challenge today - although I'm quite happy with my solution, I couldn't figure out how
to manipulate the input properly to extract the data... Since I ran out of time, I just manually
set up the data for the stacks of crates :(

Once it's set up this way however, the solution takes advantage of vector manipulation to "move"
the last item (ie crate) from one vector (ie stack) to the end of another vector. For this,
`tail()` is useful since you can manually choose how many elements you want to take from the
end of the vector (convenient for Part 2!)

## Solution


```{r result1}

input = readLines(here::here("2022","inputs","input5.txt"))

#sadly, this manual setup means my solution isn't fully reproducible...
stacks = list(c("W","B","D","N","C","F","J"),
              c("P","Z","V","Q","L","S","T"),
              c("P","Z","B","G","J","T"),
              c("D","T","L","J","Z","B","H","C"),
              c("G","V","B","J","S"),
              c("P","S","Q"),
              c("B","V","D","F","L","M","P","N"),
              c("P","S","M","F","B","D","L","R"),
              c("V","D","T","R"))

input = input[-c(1:10)]

for(instruct in input){
  
  instruct = gsub("[a-z]+ ", "", instruct) |>
    strsplit(" ") |>
    unlist() |>
    as.numeric()
  
  for(i in 1:instruct[1]){
    stacks[[instruct[3]]] = c(stacks[[instruct[3]]], tail(stacks[[instruct[2]]],1))
    stacks[[instruct[2]]] = stacks[[instruct[2]]][-length(stacks[[instruct[2]]])]
  }
  
}

lapply(stacks, function(x) tail(x,1)) |>
  unlist()

```







# Part 2

## Challenge

The only difference here is that instead of moving one crate at a time, repeated n times,
now we're moving n crates at a time. This means we can just adjust the argument in `tail()`
to move the required number of crates, then use `head()` to update the stack from which
the crates were moved.


## Solution


```{r result2}

input = readLines(here::here("2022","inputs","input5.txt"))

stacks = list(c("W","B","D","N","C","F","J"),
              c("P","Z","V","Q","L","S","T"),
              c("P","Z","B","G","J","T"),
              c("D","T","L","J","Z","B","H","C"),
              c("G","V","B","J","S"),
              c("P","S","Q"),
              c("B","V","D","F","L","M","P","N"),
              c("P","S","M","F","B","D","L","R"),
              c("V","D","T","R"))

input = input[-c(1:10)]

for(instruct in input){
  
  instruct = gsub("[a-z]+ ", "", instruct) |>
    strsplit(" ") |>
    unlist() |>
    as.numeric()
  
  stacks[[instruct[3]]] = c(stacks[[instruct[3]]], tail(stacks[[instruct[2]]],instruct[1]))
  stacks[[instruct[2]]] = head(stacks[[instruct[2]]], length(stacks[[instruct[2]]])-instruct[1])
  
}

lapply(stacks, function(x) tail(x,1)) |>
  unlist()


```

