---
title: "Advent of code 2022: day 6"
author: "Quentin Leclerc"
date: "06/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

The problem today is about looping through a vector of letters, and stopping as 
soon as 4 adjacent letters are different. My solution is therefore just a loop to do
exactly that! It's the solution I'm most proud of this year so far, the `break`
statement in the loop is nice for efficiency to avoid going through the entire
vector, and `length(unique())` is very convenient to check if all characters in
the sequence are unique.

Plus, you just need to replace 4 by 14 and 3 by 13 for it to work in Part 2!

## Solution


```{r result1}

input = strsplit(readLines(here::here("2022", "inputs", "input6.txt")), "")[[1]]

for(i in (4:length(input))){
  if(length(unique(input[(i-3):i])) == 4) break
}

i

```



# Part 2

## Solution


```{r result2}

input = strsplit(readLines(here::here("2022", "inputs", "input6.txt")), "")[[1]]

for(i in (14:length(input))){
  if(length(unique(input[(i-13):i])) == 14) break
}

i

```

