---
title: "Advent of code 2021: day 1"
author: "Quentin Leclerc"
date: "01/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2021/day/1

## Solution

Here, we have to evaluate the difference between consecutive values in a vector. Luckily, there's the built-in `diff()` function in `R` which does just that!
Then, we simply count how many differences are positive, indicating that the value has increased, and we have our answer.

Testing with the example first:

```{r test_solution1}

input = c(199,
          200,
          208,
          210,
          200,
          207,
          240,
          269,
          260,
          263)

sum(diff(input) > 0)


```

And now with the puzzle input:

```{r result1}

input = as.numeric(readLines(here::here("2021", "inputs", "input1.txt")))

sum(diff(input) > 0)


```



# Part 2

## Solution

In this second part, we first have to "group" the values in sets of 3 (ie sum each value with the two that follow it). Here I just chose to sum the vector three times, shifting the window of values to use for each vector. Once that's done, we can just apply the `diff()` method like above.

Testing with the example...

```{r test_solution2}

input = c(199,
          200,
          208,
          210,
          200,
          207,
          240,
          269,
          260,
          263)

input = input[1:c(length(input)-2)] + input[2:c(length(input)-1)] + input[3:length(input)]

sum(diff(input) > 0)

```

And now with the puzzle input:

```{r result2}

input = as.numeric(readLines(here::here("2021", "inputs", "input1.txt")))

input = input[1:c(length(input)-2)] + input[2:c(length(input)-1)] + input[3:length(input)]

sum(diff(input) > 0)

```

