---
title: "Advent of code 2021: day 13"
author: "Quentin Leclerc"
date: "13/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2021/day/13

In today's challenge, we have to "fold" a piece of paper. Part 1 just asks us to fold it once, and part 2 to fold it multiple times, so my solution to part a is directly applicable to part 2.

## Solution

The first challenge is how to work with the input. I've chosen to represent the paper as a matrix, with 0s indicating a blank space, and 1 indicating a dot. Then, to "fold" the paper, I'm splitting the matrix along the fold line, and adding up the values on one side with a flipped version of the other side. I can then just replace values greater than 1 by 1 to keep the output.

Since I knew part 2 was going to ask me to fold it multiple times, I already implemented the folding as a loop here, even though it's technically not needed.

```{r test_solution1}

library(dplyr)

input = readLines(here::here("2021", "inputs", "input13.txt"))

#separate fold instructions from initial coordinates of dots
input_fold = input[-c(1:which(input == ""))]
input = input[-c(which(input == ""):length(input))]

#reformat the input to have a nice dataset of dots coordinates
input = strsplit(input, ",") %>%
  unlist(.) %>%
  as.numeric(.) %>%
  matrix(. , ncol = 2, byrow = T)

#create our "sheet of paper", with 1s indicating dots and 0s blank spaces
paper = matrix(0, nrow = max(input[,2]+1), ncol = max(input[,1]+1))

#fill in the dots based on the coordinates
for(i in 1:nrow(input)){
  
  paper[input[i,2]+1, input[i,1]+1] = 1
  
}

#apply fold instructions
for(i in input_fold[1]){
  
  #check if we're folding along the x (columns) or y (rows) axis
  if(grepl("x", i)){
    
    #extract fold position
    #(I know this is technically unnecessary and I could just fold along the middle point but meh)
    i = as.numeric(gsub(".*=", "", i))+1
    
    #extract the first half of the matrix
    new_paper = paper[,c(1:i-1)]
    
    #reverse the second half, and add the values to the first half
    new_paper = new_paper + paper[,c(ncol(paper):(i+1))]
    
  } else {
    
    #same logic as above, but splitting by row instead of column
    
    i = as.numeric(gsub(".*=", "", i)) + 1

    new_paper = paper[c(1:i-1),]
    
    new_paper = new_paper + paper[c(nrow(paper):(i+1)),]

  }
  
  #update paper
  paper = new_paper
  #clean paper, as multiple overlapping dots (>1) are just equivalent to one dot (1)
  paper[paper > 1] = 1
  
}

#how many dots in total?
sum(paper)

```





# Part 2

## Challenge

As mentioned, in this second bit we just need to repeat the folding multiple times...

## Solution

So, my solution here is the same as above, but I noticed a glitch. I had assumed that the number of rows/columns would be the same as the maximum coordinates in my input, but that's not true! There's an extra empty row in my data... So I just manually added it here.

Also note that the patterns of 0s and 1s on our final paper display capital letters, and that those capital letters are our puzzle answer. To be honest, I can't be bothered to figure out how to code something to automatically translate this into letters, so I just opened it up, looked at it, and figured out myself what the letters were :)

```{r result2}

input = readLines(here::here("2021", "inputs", "input13.txt"))

input_fold = input[-c(1:which(input == ""))]
input = input[-c(which(input == ""):length(input))]

input = strsplit(input, ",") %>%
  unlist(.) %>%
  as.numeric(.) %>%
  matrix(. , ncol = 2, byrow = T)

#well, I only knew after trial an error that there had to be an extra blank row in the matrix...
paper = matrix(0, nrow = max(input[,2])+2, ncol = max(input[,1])+1)

for(i in 1:nrow(input)){
  
  paper[(input[i,2]+1), (input[i,1]+1)] = 1
  
}

for(i in input_fold){
  
  if(grepl("x", i)){
    
    i = as.numeric(gsub(".*=", "", i))+1
    
    new_paper = paper[,c(1:i-1)]
    
    new_paper = new_paper + paper[,c(ncol(paper):(i+1))]
    
  } else {
    
    i = as.numeric(gsub(".*=", "", i)) + 1

    new_paper = paper[c(1:i-1),]
    
    new_paper = new_paper + paper[c(nrow(paper):(i+1)),]

  }
  
  paper = new_paper
  paper[paper > 1] = 1
  
}

#figure out what the letters are myself
View(paper)

```

