---
title: "Advent of code 2020: day 4"
author: "Quentin Leclerc"
date: "04/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2021/day/4

In this puzzle, we have a set of grids, and we go through a list of numbers, each time "marking" the numbers on the grids, until one grid has a bingo (one full row or column of numbers marked.)


## Solution

The first step was to decide how to "mark" numbers, and check for a bingo. I chose to replace matching values by -1 in the grid. This way, a bingo would show up as a column or row sum of -5 (since the grids are 5*5).
Thinking about the best way to go through the grids until we find a bingo, I decided to reformat my puzzle input as a list of dataframes (each dataframe being a grid). Then, I can use `lapply()` and some custom function to replace values and check for a bingo at each step.

Full disclaimer: I struggled a fair bit to get this to work, so trying out with the example input was essential here!

```{r test_solution1}

#extract the bingo values
bingo_vals = readLines(here::here("2021", "inputs", "input4ex.txt"))[1]
bingo_vals = as.numeric(unlist(strsplit(bingo_vals, ",")))

#extract the grids
input = read.table(here::here("2021", "inputs", "input4ex.txt"),skip = 1)

#group input into 5*5 grids in list
bingo_grids = list()
j = 0

for(i in seq(1, nrow(input), 5)){
  j = j + 1
  bingo_grids[[j]] = input[i:(i+4),]
}

#function to check if a bingo happened
check_sums = function(x){
  
  c_sums = (colSums(x)==-5)
  r_sums = (rowSums(x)==-5)
  
  any(c_sums, r_sums)
  
}

#function to replace values in grids by -1 if they match the bingo value
replace_values = function(x, value){
  x[x==value] = -1
  return(x)
}

i = 0

#repeat until we get a bingo
repeat{
  
  i = i+1
  test_val = bingo_vals[i]
  
  bingo_grids = lapply(bingo_grids, replace_values, value = test_val)
  
  bingo = unlist(lapply(bingo_grids, check_sums))

  #as soon as we get a bingo, extract which grid won, and break
  if(any(bingo == TRUE)){
    good_grid = bingo_grids[[which(bingo == TRUE)]]
    break
  }
  
}

#sum all values which weren't matched (ie not equal to -1), and multiply by last bingo value
sum(good_grid[good_grid != -1]) * test_val

```

Okay, let's use the main input now:

```{r result1}

bingo_vals = readLines(here::here("2021", "inputs", "input4.txt"))[1]
bingo_vals = as.numeric(unlist(strsplit(bingo_vals, ",")))

input = read.table(here::here("2021", "inputs", "input4.txt"),skip = 1)

bingo_grids = list()
j = 0

for(i in seq(1, nrow(input), 5)){
  j = j + 1
  bingo_grids[[j]] = input[i:(i+4),]
}

i = 0

repeat{
  
  i = i+1
  test_val = bingo_vals[i]
  
  bingo_grids = lapply(bingo_grids, replace_values, value = test_val)
  
  bingo = unlist(lapply(bingo_grids, check_sums))

  if(any(bingo == TRUE)){
    good_grid = bingo_grids[[which(bingo == TRUE)]]
    break
  }
  
}

sum(good_grid[good_grid != -1]) * test_val


```





# Part 2

## Challenge

In this second part, instead of being interested in the first grid to win, we want to find the last one.

## Solution

This is a relatively easy second part thanks to the structure we've chosen to solve part 1, we just need to change the break condition to wait until all grids have a bingo (ie a column or row sum of 1). I added an extra `if` statement to catch the last grid to still not have a bingo.

Since this is straightforward, I used the main puzzle input directly:

```{r result2}

bingo_vals = readLines(here::here("2021", "inputs", "input4.txt"))[1]
bingo_vals = as.numeric(unlist(strsplit(bingo_vals, ",")))

input = read.table(here::here("2021", "inputs", "input4.txt"),skip = 1)

bingo_grids = list()
j = 0

for(i in seq(1, nrow(input), 5)){
  j = j + 1
  bingo_grids[[j]] = input[i:(i+4),]
}

i = 0

repeat{
  
  i = i+1
  test_val = bingo_vals[i]
  
  bingo_grids = lapply(bingo_grids, replace_values, value = test_val)
  
  bingo = unlist(lapply(bingo_grids, check_sums))

  #extra if statement to catch which grid is the last one to win
  if(sum(bingo == FALSE) == 1){
    good_grid = which(bingo == FALSE)
  }
  
  #break when all the grids won
  if(all(bingo == TRUE)){
    good_grid = bingo_grids[[good_grid]]
    break
  }

  
}

sum(good_grid[good_grid != -1]) * test_val


```

Yay!
