---
title: "Advent of code 2021: day 5"
author: "Quentin Leclerc"
date: "05/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2021/day/5

In this challenge, we have to use coordinates to "draw" lines, and identify how many times lines cross at least once.


## Solution

I went for a very explicit approach to solve this problem. Here, I'm creating a matrix of 0s, and at each step I'm "drawing" a line by adding 1 to the values in the matrix which are between the two coordinates for that step.

```{r test_solution1}

#a bit of reformatting to have a matrix of inputs, where each line is the set of coordinates for that step
input = readLines(here::here("2021", "inputs", "input5ex.txt"))
input = gsub(" -> ", ",", input)
input = unlist(strsplit(input, ","))
input = matrix(as.numeric(input), ncol = 4, byrow = T)

#creating the matrix to store results
result_grid = matrix(0,
                     max(c(input[,2], input[,4]))+1,
                     max(c(input[,1], input[,3]))+1)

#go through each set of coordinates
for(i in 1:nrow(input)){
  
  current_row = input[i,]
  
  #only look at vertical or horizontal lines, not all sets of coordinates
  if(current_row[1] == current_row[3] | current_row[2] == current_row[4]){
    
    #note the +1s because R indexing is from 1, not 0
    result_grid[(current_row[2]+1):(current_row[4]+1), (current_row[1]+1):(current_row[3]+1)] = result_grid[(current_row[2]+1):(current_row[4]+1), (current_row[1]+1):(current_row[3]+1)] + 1
    
  }
  
}

#points where more than 1 line has been "drawn" will have a value greater than 1
sum(result_grid > 1)

```

Works with the examples!

```{r result1}

input = readLines(here::here("2021", "inputs", "input5.txt"))
input = gsub(" -> ", ",", input)
input = unlist(strsplit(input, ","))
input = matrix(as.numeric(input), ncol = 4, byrow = T)

result_grid = matrix(0,
                     max(c(input[,2], input[,4]))+1,
                     max(c(input[,1], input[,3]))+1)

for(i in 1:nrow(input)){
  
  current_row = input[i,]
  
  if(current_row[1] == current_row[3] | current_row[2] == current_row[4]){
    
    result_grid[(current_row[2]+1):(current_row[4]+1), (current_row[1]+1):(current_row[3]+1)] = result_grid[(current_row[2]+1):(current_row[4]+1), (current_row[1]+1):(current_row[3]+1)] + 1
    
  }
  
}

sum(result_grid > 1)


```

And with the result too!




# Part 2

## Challenge

In this second step, unsurprisingly, we now also include coordinates which draw diagonal lines.

## Solution

Same idea as before, without the `if` statement (since we're looking at all coordinates now), and with a more general calculation for which values to add 1 to based on the beginning and end coordinates. This is using the "absolute" position of values in a matrix, instead of their position by coordinates.

For example, in a 4 by 4 matrix `mat`, the value `mat[2,3]` would be the same as `mat[(2-1)*4+3]`. More generally: for a matrix of `n` columns, `mat[x,y]` is equivalent to `mat[(x-1)*n+y]`.

Because the diagonals are all in steps of 1 (ie 45 degrees), we can use sequences in steps of 1 to move from a starting point to an end point. For example, if I wanted to draw a line from position `[1,3]` to `[4,6]`, the coordinates that this line will pass through will be `[2,4]`, `[3,5]`, and `[4,6]` (+1 every time).

```{r test_result2}

input = readLines(here::here("2021", "inputs", "input5ex.txt"))
input = gsub(" -> ", ",", input)
input = unlist(strsplit(input, ","))
input = matrix(as.numeric(input), ncol = 4, byrow = T)

result_grid = matrix(0,
                     max(c(input[,2], input[,4]))+1,
                     max(c(input[,1], input[,3]))+1)

for(i in 1:nrow(input)){
  
  current_row = input[i,]
  
  start_y = current_row[2]
  end_y = current_row[4]
  start_x = current_row[1] + 1
  end_x = current_row[3] + 1
  
  result_grid[c(start_y:end_y * ncol(result_grid))+c(start_x:end_x)] = result_grid[c(start_y:end_y * ncol(result_grid))+c(start_x:end_x)] + 1
    
  
}

sum(result_grid > 1)


```

Looks good with the example, so should also work with the actual input:

```{r result2}

input = readLines(here::here("2021", "inputs", "input5.txt"))
input = gsub(" -> ", ",", input)
input = unlist(strsplit(input, ","))
input = matrix(as.numeric(input), ncol = 4, byrow = T)

result_grid = matrix(0,
                     max(c(input[,2], input[,4]))+1,
                     max(c(input[,1], input[,3]))+1)

#only keep 

for(i in 1:nrow(input)){
  
  current_row = input[i,]
  
  start_y = current_row[2]
  end_y = current_row[4]
  start_x = current_row[1] + 1
  end_x = current_row[3] + 1
  
  result_grid[c(start_y:end_y * ncol(result_grid))+c(start_x:end_x)] = result_grid[c(start_y:end_y * ncol(result_grid))+c(start_x:end_x)] + 1
    
  
}

sum(result_grid > 1)


```

