---
title: "Advent of code 2021: day 9"
author: "Quentin Leclerc"
date: "09/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Challenge

https://adventofcode.com/2021/day/9

In today's puzzle, we are trying to find low points in a matrix (ie values which are small than all their neighbours).

## Solution

To solve this first section, I worked out the difference between a value and each of its four neighbours using matrices. Then, I only keep the values which always have a negative difference with their neighbours (ie which are always smaller than their neighbours).

```{r test_solution1}

library(dplyr)

#a bit of reformatting to get a nice matrix
input = readLines(here::here("2021", "inputs", "input9.txt")) %>%
  strsplit(., "") %>%
  unlist(.) %>%
  as.numeric(.) %>%
  matrix(., nrow = sqrt(length(.)), byrow = T)

#difference between values and their right neighbour
grid_1 = input - cbind(input[,-1], 10)
#difference between values and their left neighbour
grid_2 = input - cbind(10, input[,-ncol(input)])
#difference between values and their bottom neighbour
grid_3 = input - rbind(input[-1,], 10)
#difference between values and their top neighbour
grid_4 = input - rbind(10, input[-nrow(input),])

#figure out which inputs are smaller than all their neighbours, add 1, and sum
sum(input[intersect(which(grid_1<0), intersect(which(grid_2<0), intersect(which(grid_3<0), which(grid_4<0))))]+1)

```






# Part 2

## Challenge

Not gonna lie: this is messy. Here, we have to figure out basin sizes, ie the area around a low point which is surrounded by 9s (the maximum value).

## Solution

This is not a clean solution! Here, I'm using the fact that each low point belongs to one unique basin. I then loop over these low points, check their neighbours, and store these neighbours if they are not equal to 9 (ie they are part of the same basin). I then repeat this process to check the neighbours of these neighbours etc... until I see no difference between two steps, which tells me that I've identified all values which are part of the same basin.

Had to fine tune this quite a lot so using the example input was essential:

```{r test_solution2}

input = readLines(here::here("2021", "inputs", "input9ex.txt")) %>%
  strsplit(., "") %>%
  unlist(.) %>%
  as.numeric(.) %>%
  matrix(., ncol = 10, byrow = T)

#"surround" the input in 9s to make my life easier
input = cbind(9, input, 9)
input = rbind(9, input, 9)

grid_1 = input - cbind(input[,-1], 10)
grid_2 = input - cbind(10, input[,-ncol(input)])
grid_3 = input - rbind(input[-1,], 10)
grid_4 = input - rbind(10, input[-nrow(input),])

#work out which values are low points
low_points = intersect(which(grid_1<0), intersect(which(grid_2<0), intersect(which(grid_3<0), which(grid_4<0))))

basin_sizes = low_points

#replace all non-maximum values by 0 to make my life easier
input[input < 9] = 0

#loop through all low points
for(i in low_points){
  
  old_basin = c(0,0)
  new_basin = i
  
  #keep going until I've identified all values in the same basin
  while(length(new_basin) != length(old_basin)){
    
    old_basin = new_basin

    #go through all the values I have already identified as being part of the basin
    #not efficient!! Means I'm checking values multiple times
    for(j in old_basin){
      
      #extract all neighbours of that value
      surrounding_basin = c(j-1, j+1, j-nrow(input), j+nrow(input))
      #only keep neightbours that are part of the basin (ie not 9s)
      new_basin = c(new_basin, surrounding_basin[input[surrounding_basin] == 0])
      
    }
    
    #clean to remove duplicate values
    new_basin = unique(new_basin)
    
  }
  
  #store basin size for that low point
  basin_sizes[low_points == i] = length(new_basin)
  
}

prod(tail(sort(basin_sizes), 3))

```

Now it works, let's use the main input:

```{r result2}

input = readLines(here::here("2021", "inputs", "input9.txt")) %>%
  strsplit(., "") %>%
  unlist(.) %>%
  as.numeric(.) %>%
  matrix(., ncol = 100, byrow = T)

input = cbind(9, input, 9)
input = rbind(9, input, 9)

grid_1 = input - cbind(input[,-1], 10)
grid_2 = input - cbind(10, input[,-ncol(input)])
grid_3 = input - rbind(input[-1,], 10)
grid_4 = input - rbind(10, input[-nrow(input),])

low_points = intersect(which(grid_1<0), intersect(which(grid_2<0), intersect(which(grid_3<0), which(grid_4<0))))

basin_sizes = low_points

input[input < 9] = 0

for(i in low_points){
  
  old_basin = c(0,0)
  new_basin = i
  
  while(length(new_basin) != length(old_basin)){
    
    old_basin = new_basin

    for(j in old_basin){
      
      surrounding_basin = c(j-1, j+1, j-nrow(input), j+nrow(input))
      new_basin = c(new_basin, surrounding_basin[input[surrounding_basin] == 0])
      
    }
    
    new_basin = unique(new_basin)
    
  }
  
  basin_sizes[low_points == i] = length(new_basin)
  
}

prod(tail(sort(basin_sizes), 3))

```

